\input{stdinclude.tex}

\begin{document}

\mainHeaderHw{1}{Coordinate Geometry}
\subsection*{Reading}
\begin{center}
\begin{tcolorbox}[width=0.8\textwidth,colback={white},title={\textbf{Go and watch...}},colbacktitle=black,coltitle=white]
  \textcolor{black}{\url{https://www.youtube.com/watch?v=X1E7I7_r3Cw}}
\end{tcolorbox}
\end{center}

\begin{center}
\begin{tcolorbox}[width=0.8\textwidth,colback={white},title={\textbf{What's it good for?}},colbacktitle=MidnightBlue,coltitle=white]
  People use coordinate geometry for...
  \begin{itemize}
    \item Computer graphics: coordinate geometry is a compact, efficient, and simple way to store information about and manipulate
          complicated graphical data. Vector graphics programs, like Inkscape and Adobe Illustrator, store all their graphics as
          coordinate geometric objects as they take up less storage space, are easier to edit, and easier to scale and otherwise modify
          than storing objects pixel-by-pixel.
    \item Coordinate systems are used by physicists, geographers, and other people who need to describe the positions of objects in
          space relative to each other.
  \end{itemize}
\end{tcolorbox}
\end{center}

\subsection*{Questions}
[This is a sample Ministry of Education L2 assessment task for this standard.]

Develop a general method for finding the co-ordinates of the reflection of any point in the mirror line joining the points $ (0, 9/4) $ and $ (9/2, 0) $ by
completing the following steps:
\begin{questions}
  \question Reflect the point $ (4, 1) $ in the mirror line joining the points $ (0, 9/4) $ and $ (9/2, 0) $ and find the co-ordinates of the reflection;
  \question If $ (a, b) $ is any point not on the mirror line, reflect the point $ (a, b) $ in the mirror line and find the co-ordinates of the reflection.
\end{questions}

The quality of your discussion and reasoning will determine the overall grade. Show your calculations. Use appropriate mathematical statements. Clearly
communicate your strategy and method at each stage of the solution.
\end{document}
