\input{stdinclude.tex}

\begin{document}

\mainHeaderHw{18}{Graphs and Networks}
\subsection*{Reading}
\begin{center}
\begin{tcolorbox}[width=0.8\textwidth,colback={white},title={\textbf{Go and watch...}},colbacktitle=black,coltitle=white]
  \textcolor{black}{\url{https://www.youtube.com/watch?v=heKK95DAKms}}
\end{tcolorbox}
\end{center}

\begin{center}
\begin{tcolorbox}[width=0.8\textwidth,colback={white},title={\textbf{What's it good for?}},colbacktitle=MidnightBlue,coltitle=white]
  People use graphs and networks for...
  \begin{itemize}
    \item Computer science: modelling computer networks, modelling connections between objects (Google's search algorithm is based on weighted graphs).
    \item Engineering: electrical circuits can be modelled with weighted graphs.
    \item Linguistics: syntax trees and semantic networks are used when modelling language and the changes of languages over time.
    \item Physics and chemistry: graphs can be used to model crystal structures or the structures of molecules (especially in organic chemistry).
    \item Mathematics: graph theory has tight links with topology and knot theory. The four-colour theorem (the vertices of any graph that can be
          drawn with no edge crossings can be coloured with at most four colours so no adjacent vertices have the same colour) was only proved in
          the latter half of the 20th century, and was the first major proof to involve considerable work by computers.
  \end{itemize}
\end{tcolorbox}
\end{center}

\clearpage
\subsection*{Questions}
[This is a sample internal assessment task for this standard, based on those provided by the Ministry of Education.]

This assessment activity requires you to apply graphs and networks to a real-world situation. Read the entire activity before
beginning work.

The Lower North Island Logistics Company (LNILC) wants to streamline its business by redesigning its logistics network. In Resource A, a list of
town/city pairs in the current network and travel times by truck between them are given.
\begin{enumerate}
  \item Draw a weighted graph modelling the information in Resource A.
  \item Find the shortest route (in terms of travel time) between Carterton and Marton.
  \item The cost of providing a truck between two of the city pairs is directly proportional to the travel time.
    \begin{enumerate}
      \item Determine any links that will (i) be in \emph{every} minimal spanning tree, and (ii) will be in \emph{no} minimal spanning tree.
      \item Determine the smallest possible network that covers all the destinations, and costs the least to maintain.
    \end{enumerate}
  \item LNILC is considering signing a deal with KiwiRail that provides rail services between the following city pairs:
        \begin{itemize}
          \item Wellington to Palmerston North: 135 minutes
          \item Palmerston North to Whanganui: 50 minutes
          \item Whanganui to New Plymouth: 50 minutes
          \item Palmerston North to Hastings: 110 minutes
          \item Wellington to Carterton: 80 minutes
          \item Carterton to Masterton: 18 minutes
          \item Masterton to Hastings: 140 minutes
          \item Masterton to Palmerston North: 65 minutes
        \end{itemize}
        How do your answers above change, incorporating these new links?
  \item Produce a final report, incorporating your findings above, that could recommend:
        \begin{itemize}
          \item If any truck links should not be maintained;
          \item Which, if any, rail links should be included in the future network;
          \item Any other changes that could be made to improve the cost-effectiveness of the entire network, and minimise the
                time taken for packages to travel between particular major towns and cities. (For example, you may recommend that
                a particular link be kept despite not appearing in your minimum spanning tree if it significantly reduces the distance
                between two key destinations.) Resource B may be useful for this.
        \end{itemize}
\end{enumerate}
The quality of your reasoning and how well you link this context to graph and network methods will determine the overall grade. Include calculations,
diagrams or formulae, as appropriate. Clearly communicate your method using correct mathematical statements where appropriate.

\subsection*{Resource A: town-city pairs with times}
All links are bi-directional and all travel times are in minutes.

\begin{center}
\begin{tabular}{|c|c|c|}\hline
  \textbf{City 1} & \textbf{City 2} & \textbf{Travel time by road}\\\hline
  Wellington & Porirua & 27\\
  Wellington & Lower Hutt & 22\\
  Lower Hutt & Wainuiomata & 12\\
  Lower Hutt & Upper Hutt & 22\\
  Porirua & Upper Hutt & 28\\
  Porirua & Kapiti & 30\\
  Kapiti & Upper Hutt & 68\\
  Upper Hutt & Martinborough & 49\\
  Upper Hutt & Carterton & 51\\
  Carterton & Masterton & 15\\
  Martinborough & Masterton & 43\\
  Masterton & Dannevirke & 77\\
  Masterton & Palmerston North & 79\\
  Kapiti & Otaki & 20\\
  Otaki & Levin & 18\\
  Levin & Palmerston North & 40\\
  Levin & Whanganui & 74\\
  Palmerston North & Whanganui & 56\\
  Palmerston North & Fielding & 16\\
  Fielding & Marton & 25\\
  Marton & Whanganui & 29\\
  Palmerston North & Dannevirke & 54\\
  Dannevirke & Napier & 89\\
  Dannevirke & Hastings & 74\\
  Napier & Hastings & 24\\
  Whanganui & Hawera & 68\\
  Hawera & Stratford & 26\\
  Stratford & New Plymouth & 33\\
  Hawera & New Plymouth & 90\\\hline
\end{tabular}
\end{center}

\subsection*{Resource B: major towns and cities in the Lower North Island}
The following urban areas served by LNILC have populations over 10,000:
\begin{multicols}{2}
\begin{itemize}
  \item Wellington (207,900)*
  \item Napier-Hastings (133,000)*
  \item Lower Hutt (104,700)
  \item Palmerston North (85,300)
  \item New Plymouth (57,500)*
  \item Porirua (55,900)
  \item Upper Hutt (43,200)
  \item Kapiti (42,300)
  \item Whanganui (40,300)*
  \item Masterton (21,800)
  \item Levin (20,900)
  \item Fielding (16,550)
\end{itemize}
\end{multicols}
(*Urban area served by a port.)

\end{document}
