\documentclass[a4paper]{exam}

\usepackage{siunitx}
\DeclareSIUnit{\revolution}{rev}
\DeclareSIUnit{\rpm}{\revolution\per\minute}
\DeclareSIUnit{\lightyear}{ly}

\begin{document}
  \section*{L3 Physics: Questions for 2.1-2.4}
  Majority taken from Knight chap. 8, 12.
  \begin{questions}
    \question A \SI{1350}{\kilo\gram} car drives around a flat \SI{150}{\metre} diameter circular track at \SI{20}{\metre\per\second}.
              What is the magnitude and direction of the net force on the car? What causes this force?
    \question A car drives over the top of a circular-section hill with a radius of \SI{50}{\metre}. Draw a force diagram, and calculate
              the maximum speed the car can drive without taking flight.
    \question A new car is tested on a \SI{200}{\metre} diameter circular track. If the car speeds up at a steady \SI{1.5}{\metre\per\second},
              how long after starting is its centripetal acceleration equal to the tangential acceleration?
    \question A \SI{300}{\gram} ball and a \SI{600}{\gram} ball are connected by a \SI{40}{\centi\metre} rigid massless rod. The structure rotates
              about its centre of mass at \SI{100}{\rpm}. What is its rotational kinetic energy?
    \question A ball of radius $ R $ is placed at aa height of \SI{30}{\centi\metre} on a \SI{15}{\degree} slope. It is released and rolls
              without slipping to the bottom. From what height should a circular hoop of radius $ R $ be released in order to equal the
              ball's speed at the bottom?
    \question A steel beam, unmelted by jet fuel, is \SI{5}{\metre} long and has a mass of \SI{400}{\kilo\gram}. It extends horizontally
              from the point to which it has been bolted to the framework of a building. A construction worker of mass \SI{100}{\kg}, unworried
              about health and safety regulations, stands on the end of the beam. What is the total torque about the bolt due to the worker
              and the beam?
    \question A \SI{2.0}{\kilo\gram}, \SI{30}{\centi\metre} diameter disc is spinning at \SI{300}{\rpm}. How much friction force must the brake
              apply to the rim to bring the disc to a halt in \SI{3.0}{\second}?
    \question A \SI{200}{\gram} toy car is placed on a narrow \SI{60}{\centi\metre} diameter track (mass of track ring \SI{1.0}{\kilo\gram},
              massless interior) that is free to rotate on a frictionless vertical axis. When the car is started, it soon reaches a steady
              speed of \SI{0.75}{\metre\per\second} relative to the track. At this point, what is the track's angular velocity in rpm?
    \question {[Difficult]} If a vertical cylinder of some liquid rotates about its axis, the surface forms a smooth curve. Assuming that the water
              moves as a unit (i.e. all the water moves with the same angular velocity), show that the shape of the surface is a parabola described
              by the equation
              \begin{equation}
                z = \frac{\omega^2}{2g} r^2.
              \end{equation}
              [Hint: Each particle of water on the surface is subject to only two forces: gravity, and the normal force due to the water underneath
              it. The normal force, as always, acts perpendicular to the surface.]
  \end{questions}
\end{document}
