\input{stdinclude.tex}

\begin{document}

\mainHeaderDiff{1}{The Derivative}

Last year we met the derivative. Let us recall that the derivative of a function is, in some sense:
\begin{itemize}
  \item A way to find the slope of another function.
  \item A function giving a rate of change.
  \item A way to find maxima and minima of a function.
\end{itemize}
The value of the derivative of a function at a point is the gradient/slope/rate of change of the function at that
point. Over the next few sections, we will make this rather vague notion precise, allowing us to transform geometric questions
about slope, curviness and speed into questions that can be solved via algebraic manipulation.

If $ f $ is a function, then the derivative is a function that takes some point $ x $ and gives us the slope of $ f $ at $ x $. For this
function, we write $ f' $ (read \emph{eff prime}), so the slope of $ f $ at $ x $ is denoted by $ f'(x) $. If we want to be explicit about the names
of the quantities we are relating, then if $ y = f(x) $ we write $ f'(x) = \od{y}{x} $ (read \emph{dee y over dee x}); in this notation,
the slope of the graph at $ x = x_0 $ is written as $ \eval{\od{y}{x}}_{x = x_0} $ (but I tend to avoid this latter notation).

If a function does not have a well-defined derivative at some point, it is said to be non-differentiable at that point.

\begin{rem}
  Note that this isn't a proper definition for the derivative: all we've done is say that the derivative `is the slope' of a function. However,
  we only have a definition of \emph{slope} for linear functions. Our solution to this will be, next week, a definition of derivatives that doesn't
  mention slope explicitly but generalises the definition of slope.
\end{rem}

\begin{ex}
  Consider a function defined by $ y = 2x + 1 $. At every point, the slope of the graph of this function is 2; so
  the value of the derivative of this function is 2 at every point. (Hence $ \od{y}{x} = 2 $.)
\end{ex}

\begin{ex}
  The speed of a particle at any time is the rate of change of the displacement of the particle at that time;
  this can also be viewed as the slope of the displacement-time graph of the particle. Hence the speed of the
  particle is the derivative of the displacement.
\end{ex}

\begin{ex}
  In the diagram, the derivative of the red functions is shown as the blue function. Note
  that the derivative of the function depends only upon the shape, not the $ y$-shift. Also,
  see that the derivative is positive as the function increases and is negative when the
  function is decreasing. The derivative is zero exactly where the function is `flat'.

  \begin{center}
    \fbox{\begin{tikzpicture}
      \begin{axis}[
        axis lines = center,
        xlabel = $ x $,
        ylabel = $ y $
      ]
        \addplot[domain = -5:5, color = red] {(x - 2)*(x)*(x + 4)};
        \addplot[domain = -5:5, color = red] {(x - 2)*(x)*(x + 4) - 30};
        \addplot[domain = -5:5, color = red] {(x - 2)*(x)*(x + 4) + 20};
        \addplot[domain = -5:5, color = blue] {3*x^2 + 4*x - 8};
      \end{axis}
    \end{tikzpicture}}
  \end{center}
\end{ex}

\clearpage
\subsection*{Questions}
\begin{questions}%\item
  \questioA Why does the derivative not depend on the $ y$-shift of a function?
  \questioA Let $ f $ be a function. Describe the difference between $ f $, $ f' $, $ f(57) $, and $ f'(57) $.
  \questioA Draw the graph of the derivative of each graphed function.
    \begin{multicols}{2}
    \begin{parts}
      \part
        \fbox{\begin{tikzpicture}
          \begin{axis}[
            axis lines = center,
            xlabel = $ x $,
            ylabel = $ y $
          ]
            \addplot[domain = -5:5, color = YellowOrange, samples=100] {sin(deg(x))};
          \end{axis}
        \end{tikzpicture}}
      \part
        \fbox{\begin{tikzpicture}
          \begin{axis}[
            axis lines = center,
            xlabel = $ x $,
            ylabel = $ y $
          ]
            \addplot[domain = -5:5, color = green] {-x^2 + 2};
          \end{axis}
        \end{tikzpicture}}
      \part
        \fbox{\begin{tikzpicture}
          \begin{axis}[
            axis lines = center,
            xlabel = $ x $,
            ylabel = $ y $
          ]
            \addplot[domain = -5:5, color = red, samples=100] {0.02*x^5 - 0.5*x^3 + x^2 - sin(x)};
          \end{axis}
        \end{tikzpicture}}
      \part
        \fbox{\begin{tikzpicture}
          \begin{axis}[
            axis lines = center,
            xlabel = $ x $,
            ylabel = $ y $
          ]
            \addplot[domain = -5:5, color = blue, samples=100] {ln(x^2)};
          \end{axis}
        \end{tikzpicture}}
    \end{parts}
    \end{multicols}
  \questioA Describe several ways in which a function $ f $ can fail to be differentiable at a point $ x $, illustrating your examples with sketches.
  \questioA Consider each of these functions in turn. Where is the derivative of each (i) negative, (ii) positive, (iii) zero, and (iv) undefined?
    \begin{parts}
      \part $ x \mapsto x^2 $
      \part $ x \mapsto \sin x $
      \part $ x \mapsto \tan x $
    \end{parts}
  \questioA Describe the derivative of the function $ x \mapsto \tan^{-1} x $.
  \questioA Write down the slope of $ y = mx + c $. Hence give $ \od{y}{x} $.
  \questioA If $ f' $ is the derivative of some function $ f $, describe the meaning of $ (f')' $ in terms of rate of change (to save space, from now on we
            will write $ f'' $ for this second derivative).
  \questioA Let $ f $ be a function. Suppose that it is known that $ f'(3) = 9 $, and $ f(3) = 6 $.
    \begin{parts}
      \part What does the graph of $ y = f(x) $ look like around $ x = 3 $?
      \part Give the equation of the tangent line to $ f(x) $ at $ x = 3 $.
    \end{parts}
  \questioM The function $ \mathrm{floor} $ maps any number $ x $ to the greatest integer less than or equal to $ x $ (so $ \mathrm{floor}(\pi) = 3 $, for
            example). Draw a graph of $ y = \mathrm{floor}(x) $. Where is $ \mathrm{floor} $ differentiable?
  \questioM When a hot water tap is turned on, the temperature $ T $ of the water depends on how long
            the water has been running.
    \begin{parts}
      \part Sketch a possible graph of $ T $ as a function of the time $ t $ that the tap has been running.
      \part Describe how the rate of change of $ T $ with respect to $ t $ varies as $ t $ increases.
      \part Sketch a graph of the derivative of $ T $ with respect to $ t $.
    \end{parts}
  \questioM The rate of change of a population at a time $ t $ is directly proportional to the
            population $ P(t) $ at that time, such that $ \od{P}{t} = P $. Draw a graph of the
            population over time if $ P(0) \approx 1000 $.
  \questioM Consider a continuous real-valued function $ f $ which has $ n $ roots (all of multiplicity one) in some
            interval $ I $. How many roots must its derivative have in $ I $?
  \questioM If a function is periodic, with a period of $ T $, what can you say about its derivative?
  \questioM Consider an ellipse, $ \frac{x^2}{a^2} + \frac{y^2}{b^2} = 1 $. This is not a function: since both $ (0, b) $ and $ (0, -b) $ are
            members of the function, it fails the vertical line test. However, it would be nice to reason about its
            rate of change \textit{as if it were} a function. Describe the slope of the ellipse as a particle traces the curve in
            an anticlockwise direction at a constant rate.
  \questioM The number of bacteria after $ t $ hours in a controlled laboratory experiment is $ n = f(t) $.
    \begin{parts}
      \part What is the meaning of the derivative $ f'(5) $? What are its units?
      \part Suppose that there is an unlimited amount of space and nutrients. Which would you
            expect to be larger, $ f'(5) $ or $ f'(10) $? If the supply of nutrients is limited
            does your answer change?
    \end{parts}
\end{questions}
\end{document}
