\input{stdinclude.tex}
\begin{document}

\mainHeaderDiff{4}{The Chain Rule}
\begin{center}
  \fbox{\begin{tikzpicture}
    \begin{axis}[
      axis lines = center,
      xlabel = $ x $,
      ylabel = $ y $,
      vasymptote = 1.2533141373155001
    ]
      \addplot[domain = 0:6.28, color = red, samples=200] {sin(deg(x^2))};
    \end{axis}
  \end{tikzpicture}}
\end{center}
Consider the function $ x \mapsto \sin (x^2) $. This function is made up of two
functions, applied one after the other:
\begin{displaymath}
  x \xmapsto{f} x^2 \xmapsto{g} \sin(x^2).
\end{displaymath}
We often notate this \textit{function composition} as $ g \circ f $ (note that we
evaluate from the right, so $ (g \circ f)(x) = g(f(x)) $).

Obviously the derivative of $ \sin(x^2) $ is not just $ \cos(2x) $, since the former
has a stationary point at $ x = \sqrt{\dfrac{\pi}{2}} $ but $ \cos(\sqrt{2\pi}) \neq 0 $. This
shows us that, in general, the derivative of a function composition is not simply the composition
of the derivatives.

In fact, it turns out that the derivative of $ f \circ g $ is $ g' (f' \circ g) $; in other words,
\begin{displaymath}
  \od{}{x} f(g(x)) = g'(x) f'(g(x)).
\end{displaymath}
This is known as the \textit{chain rule}, since we are "chaining" together functions.

\begin{ex}
  The correct derivative of $ \sin(x^2) $ is $ 2x \cos(x^2) $.
\end{ex}

\begin{ex}
  If $ f(r) = \sqrt{r^2 - 3} $, then $ f'(r) = 2r \frac{1}{2} \left(r^2 - 3\right)^{-1/2} = \dfrac{r}{\sqrt{r^2 - 3}} $.
\end{ex}

\begin{ex}
  If $ g(x) = \sin((\sin^7 x^7 + 1)^7) $, then we compute:
  \begin{align*}
    g(x)  &= \sin \left( \left[ \left( \sin x^7 \right)^7 + 1 \right]^7 \right)\\
    g'(x) &= 7x^6 \cdot \cos x^7 \cdot 7\left(\sin x^7\right)^6 \cdot 7\left[\left(\sin x^7\right)^7 + 1\right]
                  \cdot \cos \left( \left[ \left( \sin x^7 \right)^7 + 1 \right]^7 \right)
  \end{align*}
  This result can probably be simplified, however the point is to evaluate the derivative chain from inside to outside in a systematic fashion.
\end{ex}

\begin{ex}
  One of the main uses of the chain rule is in related rates problems. For example, consider
  a disc of radius $ r $. The area of this disc is given by $ A = \pi r^2 $, and so $ \od{A}{r} = 2\pi r $.
  But what if $ r $ is itself changing with respect to time, say at a rate of $ \SI{3}{\metre\per\second} $? Then
  $ \od{A}{t} = \od{r}{t} (2\pi r) = 6\pi r = 6\pi (r_0 + 3t) $ (where $ r_0 $ is the radius at $ t = 0 $).
\end{ex}

\clearpage
\subsection*{Questions}
\begin{questions}
   \questioA Identify the inner and outer functions, but do not attempt to differentiate.
     \begin{parts}
       \part $ \sqrt{\sin x} $
       \part $ \sin \cos \tan x $
       \part $ (2x + 3)^{17} $
       \part $ 97(x + 2)^2 $
       \part $ \ln \sin x $
       \part $ \frac{1}{\sqrt{23x - x^2}} $
     \end{parts}
   \questioA Differentiate with respect to $ t $:
     \begin{parts}
       \part $ (2t + 3)^{3000} $
       \part $ \sin \ln t $
       \part $ \sqrt{t^3 + 10t^2 + 3} $
       \part $ \csc e^t $
       \part $ \sin^3 t + 14\ln (3t) $
       \part $ \sin \sin \sin t $
       \part $ \cot (t + \sec t) $
       \part $ \sin^2 ((t + \sin t)^2) $
       \part $ \ln \sqrt{t + 9} $
       \part $ \sqrt{t} + \frac{1}{\sqrt[3]{t^4}} $
       \part $ e^{\sec (t^2)} $
       \part $ \sin \sqrt{t + \tan t} $
     \end{parts}
  \questioA The derivative of a function is $ 2 \cos 2x $. What could the original function be?
  \questioM Differentiate $ y = \sin^2 x + \cos^2 x $, and hence prove that $ \sin^2 x + \cos^2 x = 1 $.
  \questioA Suppose that the displacement of a particle on a vibrating spring is given by $ x(t) =  5 + \frac{1}{8} \sin(5\pi t) $,
            where $ x $ is measured in centimetres and $ t $ in seconds.
    \begin{parts}
      \part Find the velocity of the particle at time $ t $.
      \part At which times is the particle momentarily stationary?
    \end{parts}
  \question The volume of a spherical balloon at a time $ t $ is given by $ V(t) $, and its radius is given by $ r(t) $.
    \begin{parts}
      \parA What do the derivatives $ \od{V}{t} $ and $ \od{V}{r} $ represent?
      \parM The volume of a sphere of radius $ r $ is $ V_\text{sphere} = \frac{4}{3} \pi r^3 $. Find $ \od{V}{t} $ in terms of $ \od{r}{t} $.
    \end{parts}
  \questioM If $ F(x) = f(3f(4f(x))) $, where $ f(0) = 0 $ and $ f'(0) = 2 $, find $ F'(0) $.
  \questioA Suppose $ f(x) = g(x + g(a)) $ for some differentiable function $ g $ and constant $ a $. Find $ f'(x) $.
  \question The depth of water at the end of a jetty in a harbour varies with time due to the tides. The depth
            of the water is given by the formula
            \begin{displaymath}
              W = 4.5 - 1.2 \cos \frac{\pi t}{6}
            \end{displaymath}
            where $ W $ is the water depth in metres, and $ t $ is the time in hours after midnight.
    \begin{parts}
      \parA What is the rate of change of water depth 5 hours after midnight?
      \parM When is the first time after $ t = 0 $ that the tide changes direction?
      \parE At that time, is the water changing from rising to falling or from falling to rising?
    \end{parts}
  \question Consider the function $ \psi $ given by
            \begin{displaymath}
              \psi(t) = \frac{e^t + e^{-t}}{2}
            \end{displaymath}
    \begin{parts}
      \parA Compute $ \psi' $.
      \parM Find $ \psi^{(2017)} $ (the 2017th derivative of $ \psi $).
      \parA Show that $ \psi $ satisfies the differential equation $ \od{}{t}(\psi + \psi') = \psi + \psi' $.
    \end{parts}
  \question The force $ F $ (in newtons) acting at an angle $ \theta $ with the horizontal that is needed
            to drag a mass of $ W $ kilograms along a horizontal surface at a constant velocity is given by
            \begin{displaymath}
              F = \frac{\mu W}{\cos \theta + \mu \sin \theta}
            \end{displaymath}
            where $ \mu $ is the coefficient of static friction (a constant).
    \begin{parts}
      \parA If $ W = \SI{200}{\kilo\gram} $ and $ \mu = 0.2 $, find $ \od{F}{\theta} $ when $ \theta = \frac{\pi}{6}\thinspace\si{\radian} $.
      \parM Suppose now that $ \theta $ is a function of time, so that $ \od{\theta}{t} = \SI{0.5}{\radian / \second} $. Find $ \od{F}{t} $.
    \end{parts}
  \questioE Prove that
            \begin{displaymath}
              \od{}{x} \abs{x} = \frac{x}{\abs{x}}.
            \end{displaymath}
            [\textit{Hint: Write $ \abs{x} = \sqrt{x^2} $.}]
\end{questions}
\end{document}
