\input{stdinclude.tex}
\begin{document}

\mainHeaderDiff{5}{The Product and Quotient Rules}
Two weeks ago we saw that the derivative of a product is not simply the product of the derivatives;
for example, take $ (x)(x) $. If we differentiate each term and multiply, we obtain 1; however,
the derivative of $ x^2 $ is (of course) $ 2x $. Ensure you understand why this is a counterexample
to the naive rule $ (fg)' = f'g' $ before continuing.

Suppose $ f $ and $ g $ are functions; then the \textit{real} product rule is
\begin{displaymath}
  (fg)' = gf' + fg'.
\end{displaymath}

\begin{proof}
  We simply apply the limit laws to the definition of the derivative and use a little trick (note that the expression highlighted
  in purple is exactly zero).
  \begin{align*}
    (fg)'(x) &= \lim_{h\to 0} \frac{f(x + h)g(x + h) - f(x) g(x)}{h}\\
             &= \lim_{h \to 0} \frac{f(x + h) g(x + h) \textcolor{RedViolet}{\mathbf{ {} - f(x + h)g(x) + f(x + h)g(x)}} - f(x)g(x)}{h}\\
             &= \lim_{h \to 0} \frac{f(x + h)[g(x + h) - g(x)]}{h} + \lim_{h\to 0} \frac{[f(x + h) - f(x)]g(x)}{h}\\
             &= f(x) \lim_{h \to 0} \frac{g(x + h) - g(x)}{h} + g(x) \lim_{h\to 0} \frac{f(x + h) - f(x)}{h} = f(x) g'(x) + g(x) f'(x).
  \end{align*}
\end{proof}

\begin{ex}
  Consider $ y = 2t \sin t $. Then $ \od{y}{t} = 2 \sin t + 2t \cos t $.
\end{ex}

We can also write a rule for the derivative of a quotient of functions. You will be asked to
prove it as an exercise, using the product rule.
\begin{displaymath}
  \left( \frac{f}{g} \right)' = \frac{gf' - fg'}{g^2}
\end{displaymath}

\begin{ex}
  Consider $ f(s) = \sqrt{\frac{s^2 + 1}{s^2 + 4}} $. Then
  \begin{align*}
    f'(s) &= \od{}{s} \left[ \frac{s^1 + 1}{s^2 + 4} \right] \cdot \frac{1}{2\sqrt{\frac{s^2 + 1}{s^2 + 4}}}\\
          &= \frac{(s^2 + 4)(2s) - (s^2 + 1)(2s)}{(s^2 + 4)^2} \cdot \frac{\sqrt{s^2 + 4}}{2\sqrt{s^2 + 1}}\\
          &= \frac{6s}{(s^2 + 4)^{\frac{3}{2}} (s^2 + 1)^{\frac{1}{2}}}
  \end{align*}
\end{ex}

\subsection*{Questions}
\begin{questions}
  \questioA In each case, find $ \od{y}{t} $.
    \begin{multicols}{2}
    \begin{parts}
      \part $ y = \left(3 + 2t^2\right)^4 $
      \part $ y = \frac{t^3}{\ln t} $
      \part $ y = t\sqrt{t} $
      \part $ y = 2t \sin t - (t^2 - 2) \cos t $
      \part $ y = \frac{t}{\sqrt{a^2 - t^2}} $ ($ a $ constant)
      \part $ y = \frac{1}{8} t^8 \left(1 - t^2\right)^{-4} $
      \part $ y = e^t \ln t $
      \part $ y = \log \left[1 + \frac{t^2 + 3t + 17}{t^{17}}\right] $
      \part $ y = \sin \left[e^{\tan t} \ln \tan t\right] $
      \part $ y = \frac{3t - 2}{\sqrt{2t + 1}} $
      \part $ y = \frac{\sec 2t}{1 + \tan 2t} $
      \part $ y = \frac{(t - 1)(t - 4)}{(t - 2)(t - 3)} $
      \part $ y = t \sin^2(\cos \sqrt{\sin \pi t}) $
      \part $ y = \sqrt[5]{t \tan t} $
      \part $ y = \dfrac{(t + \lambda)^4}{t^4 + \lambda^4} $
    \end{parts}
    \end{multicols}
  \questioA If $ f(x) = e^{-x} $, find $ f(0) + xf'(0) $.
  \questioM Suppose $ f $ and $ g $ are functions ($ g $ not the zero function). Write $ \frac{f}{g} = fg^{-1} $ and
            prove the quotient rule from the product rule.
  \questioM Show that if $ f $, $ g $, and $ h $ are functions then $ (fgh)' = f'gh + fg'h + fgh' $.
  \questioE Suppose $ f(x) = f(-x) $ for all $ x $ in the domain of $ f $. Prove that $ f'(x) = -f'(-x) $ for all $ x $
            in the domain of $ f'(x) $.
  \questioE Consider the function defined by $ f(x) = x^x $.
    \begin{parts}
      \part Rewrite $ f $ in the form $ f(x) = e^{x \ln x} $, and hence find $ f'(x) $.
      \part Find $ \od{y}{t} $ if $ y = (t^2 + 3)^{(t^2 + 3)} $.
    \end{parts}
  \questioE A circle that closely fits points on a local section of a curve can be drawn for any continuous curve. The
            radius of curvature of the curve is defined as the radius of the approximating circle, which changes as
            we move around the curve.
            \begin{displaymath}
              \textrm{radius of curvature} = \frac{\left[1 +  \left(\od{y}{x}\right)^2 \right]^{\frac{3}{2}}}{\abs{\od[2]{y}{x}}}
            \end{displaymath}
            Find the radius of curvature of the curve $ y = e^{-x} \sin x $ at the point $ (0, 0) $.
  \questioM Show that $ y = xe^{-x} $ satisfies the differential equation $ xy' = (1-x)y $.
  \questioM If $ y = \ln \frac{1 + \sqrt{\sin x}}{1 - \sqrt{\sin x}} $, find $ y'' $.
  \questioM Find the equation of the tangent line to the graph of $ y = \ln \cos \frac{x - 1}{x} $ at
            the point $(1, 0)$.
  \questioM Show that $ y = (1 + x + \ln x)^{-1} $ satisfies the differential equation $ xy' = y(y \ln x - 1) $.
  \questioE Find the angle at which $ y = x^2 \ln [(x - 2)^2] $ cuts the $ x$-axis at the point $ (0,0) $.
  \questioM When $ x = 0 $, is the curve $ y = (x + 20)^2 (2x^2 - 3)^6 - \ln \sin (x - \frac{\pi}{2}) $ concave up or concave down?
  \questioM If $ y = \frac{e^x}{\sin x} $, show that $ \od{y}{x} = y(1 - \cot x) $.
  \questioE Find $ f'(x + 3) $ if $ f(x + 3) = (x + 5)^7 $.
  \questioE The number $ a $ is called a \textbf{double root} of some polynomial function $ f $ if $ f(x) = (x - a)^2g(x) $ for
            some polynomial $ g $. Prove that $ a $ is a double root of $ f $ if and only if $ a $ is a root of both $ f $ and $ f' $.
  \questioE Show that there is no function of the form
            \begin{displaymath}
              f(x) = a_n x^n + a_{n - 1} x^{n - 1} \cdots + a_1 x + a_0 + \frac{b_1}{x} + \cdots + \frac{b_m}{x^m}
            \end{displaymath}
            such that $ f'(x) = 1/x $.
  \questioS Let's try to generalise the product rule for $ n$th derivatives of products. Recall that the notation $ f^{(n)}(x) $ means the $ n$th derivative
            of $ f $ with respect to $ x $. Assume in each case that the relevant derivatives exist.
    \begin{parts}
      \part Find a formula for $ (fg)'' = (fg)^{(2)} $ in terms of the derivatives of $ f $ and $ g $.
      \part What about $ (fg)^{(3)} $?
      \part If you can, prove Leibniz' formula for the $ n$th derivative of a product using \textit{induction}:
            \begin{displaymath}
              (fg)^{(n)}(x) = \sum_{k = 0}^n \binom{n}{k} f^{(k)}(x) g^{(n - k)}(x)
            \end{displaymath}
    \end{parts}
\end{questions}
\end{document}
