\documentclass[a4paper,leqno]{article}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage[inline]{enumitem}

\usepackage{siunitx}
\usepackage{multirow}
\usepackage{subcaption}


\usepackage[english]{babel}
\usepackage[autostyle, english=british]{csquotes}
\MakeOuterQuote{"}

\usepackage{commath}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}

\usepackage{skull}

\usepackage{pgfplots}
\pgfplotsset{compat=1.11}
\usepgfplotslibrary{fillbetween}
\usetikzlibrary{patterns}

\usepackage{hyperref}

\usepackage[margin=1in]{geometry}
\usepackage{changepage}
\usepackage{titlesec}
\titleformat{\section}{\normalfont\Large\bfseries\centering}{Section~\thesection:}{1em}{}

\def\signed #1{{\leavevmode\unskip\nobreak\hfil\penalty50\hskip2em
  \hbox{}\nobreak\hfil(#1)%
  \parfillskip=0pt \finalhyphendemerits=0 \endgraf}}
\newsavebox\mybox
\newenvironment{aquote}[1]
  {\savebox\mybox{#1}\begin{quote}}
  {\signed{\usebox\mybox}\end{quote}}

% Augmented matrices.
\makeatletter
\renewcommand*\env@matrix[1][*\c@MaxMatrixCols c]{%
  \hskip -\arraycolsep
  \let\@ifnextchar\new@ifnextchar
  \array{#1}}
\makeatother

%--------grstep
% For denoting a Gauss' reduction step.
% Use as: \grstep{\rho_1+\rho_3} or \grstep[2\rho_5 \\ 3\rho_6]{\rho_1+\rho_3}
\newcommand{\grstep}[2][\relax]{%
   \ensuremath{\mathrel{
       {\mathop{\longrightarrow}\limits^{#2\mathstrut}_{
                                     \begin{subarray}{l} #1 \end{subarray}}}}}}
\newcommand{\swap}{\leftrightarrow}


\swapnumbers
\numberwithin{equation}{section}
\newtheorem{thm}[equation]{Theorem}
\newtheorem{lem}[equation]{Lemma}
\newtheorem{cor}[equation]{Corollary}
\newtheorem{prp}[equation]{Proposition}
\theoremstyle{definition}
\newtheorem{defn}[equation]{Definition}
\newtheorem{ex}[equation]{Example}
\newtheorem{exercise}[equation]{Exercise}
\newtheorem{alg}[equation]{Algorithm}
\theoremstyle{remark}
\newtheorem{rem}[equation]{Remark}
\newtheorem{warning}[equation]{Warning $\skull$}

\newcommand{\df}[1]{\textbf{#1}}
\newcommand{\T}{\mathrm{T}}
\newcommand{\F}{\mathrm{F}}
\newcommand{\IndSet}{\mathbf{I}}

\DeclareMathOperator{\Map}{Map}
\DeclareMathOperator{\Sur}{Sur}
\DeclareMathOperator{\Inj}{Inj}
\DeclareMathOperator{\Bij}{Bij}

\title{Down the River: Foundations for Scholarship}
\author{Alex Elzenaar}
\date{\today}

\begin{document}
\maketitle
\setcounter{tocdepth}{1}
\tableofcontents
\section*{Preface}
This set of notes is an introduction to the shallower foundations of mathematics, for those students who
are attempting the New Zealand Scholarship examination in Calculus. We broadly cover the following topics,
but the point is to learn the methods, not the results.
\begin{itemize}
  \item A bit of set theory.
  \item A bit of rigorous combinatorics.
  \item A bit of number theory.
  \item A bit of geometry.
\end{itemize}

I assume knowledge of the breadth of Level 2, but not any material from Level 3. As a guideline, I would
expect any student attempting Scholarship to come into Level 3 with the following skills:
\begin{itemize}
  \item Comfort with basic algebraic manipulations.
  \item Understanding of exponential and logarithmic functions.
  \item Understanding of quadratic equations (especially with the link between parabolae and quadratics, and factoring).
  \item Knowledge of the links between functions and their graphs (in general; particularly, being able to give some kind
        of definition and knowledge of some basic examples).
  \item Comfort with the geometry underlying the derivative (slopes of lines, average slopes of curves, and the ability
        to apply the derivative in geometric situations like finding turning points). [This skill is not explicitly required for these notes.]
\end{itemize}

\titleformat{\section}{\clearpage\titlerule[0.8pt]\vspace{0.5ex}\normalfont\Large\bfseries\centering}{Section~\thesection:}{1em}{}[{\titlerule[0.8pt]}]
\let\oldsection\section
\renewcommand\section{\clearpage\oldsection}
\section{Set Notation}
\subsection{Defining Sets}
We begin by quoting from Paul Halmos' classic book, \textit{Naive Set Theory}.
\begin{aquote}{Halmos}
  A pack of wolves, a bunch of grapes, or a flock of pigeons are all examples of sets of things. The mathematical concept of a set can
  be used as the foundation for all known mathematics...
\end{aquote}
Like Halmos, we will avoid an exact definition of sets so that we do not need to deal with the logical issues that come
with it (does the set of all sets that do not contain themselves contain itself?) --- for us, a set is simply a collection
of objects.

We can write down a set by placing the names of the objects within the set (the \df{elements} of the set) inside curly brackets; for example,
\begin{equation}
  S = \{1, 2, 3\}
\end{equation}
is a set consisting of three elements. If $ x $ is an element of a set $ S $, then we write $ x \in S $ ($ x $ is in $ S $); otherwise,
we can write $ x \notin S $.

There are several `standard' sets:
\begin{itemize}
  \item $ \mathbb{N} = \{1, 2, 3, ...\} $ is the set of natural numbers.
  \item $ \mathbb{N}_0 = \{0, 1, 2, ...\} $ is the set of natural numbers together with zero.
  \item $ \mathbb{Z} = \{..., -2, -1, 0, 1, 2, ...\} $ is the set of integers.\footnote{The letter $ Z $ comes from the German \textit{Zahlen} (number).}
  \item $ \mathbb{Q} $ is the set of rational numbers (that is, all numbers $ x $ which can be written in the form $ a/b $ for $ a $ and $ b $ integers).
  \item $ \mathbb{R} $ is the set of real numbers (that is, all possible lengths that can be measured along an infinite line\footnote{There are more rigorous
        definitions of the real numbers than this, but we do not need to worry too much about defining them properly.}).
  \item $ \emptyset $ is the set containing no elements.
\end{itemize}

If we construct a set using a rule, then we have another notation: if $ E $ is the set containing all even numbers, then we can write
\begin{equation}
  E = \{ n : n\text{ is an even number}\}.
\end{equation}
This notation is called \df{set-builder notation}; if $ S = \{ x \in \mathcal{U} : P(x) \text{ is true}\} $, then we read it
as `$ S $ is the set of all $ x $ in $ \mathcal{U} $ such that $ P(x) $ is true'.

If $ S $ and $ T $ are sets such that every element of $ T $ is also an element of $ S $, then we say that $ T $ is a \df{subset} of $ S $ (or
that $ S $ \df{includes} $ T $), and write $ T \subseteq S $. In particular, we have $ \mathbb{N} \subseteq \mathbb{Z} \subseteq \mathbb{Q} \subseteq \mathbb{R} $.
\begin{exercise}
  Let $ X $ be some set. Prove that $ \emptyset \subseteq X $, and that $ X \subseteq X $.
\end{exercise}

\subsection{Set Operations}
If $ S $ and $ T $ are two sets, then their \df{intersection} is the set
\begin{equation}
  S \cap T = \{ x : x \in S \text{ and } x \in T \}
\end{equation}
and their \df{union} is the set
\begin{equation}
  S \cup T = \{ x : x \in S \text{ or } x \in T \}.
\end{equation}

\begin{ex}
  If $ S = \{2, 4, 6, 8, 10, 12\} $ and $ T = \{1, 4, 9, 16\} $, then $ S \cup T = \{1,2,3,6,8,9,10,12,16\} $ and $ S \cap T = \{4\} $.
\end{ex}

\begin{exercise}[Distributivity]
  Show that, if $ R $, $ S $, and $ T $ are sets, then
  \begin{enumerate}
    \item $ R \cup (S \cap T) = (R \cup S) \cap (R \cup T) $, and
    \item $ R \cap (S \cup T) = (R \cap S) \cup (R \cap T) $.
  \end{enumerate}
  Draw Venn diagrams to show each configuration.
\end{exercise}

Now, suppose $ S $ is a subset of $ \mathcal{U} $. The \df{compliment} of $ S $ in $ \mathcal{U} $ is the set of
all elements of $ \mathcal{U} $ that are not in $ S $; in other words,
\begin{equation}
  \mathcal{U} - S = \{ x \in \mathcal{U} : c \notin S\}.
\end{equation}
If the set $ \mathcal{U} $ is clear from context, we can just write $ S^C $ or $ \overline{S} $ in place of $ \mathcal{U} - S $.

\begin{exercise}[de Morgan's laws]
  Show that, if $ R $ and $ S $ are subsets of $ \mathcal{U} $, then
  \begin{enumerate}
    \item $ \overline{R \cup S} = \overline{R} \cap \overline{S} $, and
    \item $ \overline{R \cap S} = \overline{R} \cup \overline{S} $.
  \end{enumerate}
  Draw Venn diagrams to show each configuration.
\end{exercise}

\subsection{Functions}
We also define the \df{Cartesian product} of $ S $ and $ T $ by
\begin{equation}
  S \times T = \{(s,t) : s \in S, t \in T\}
\end{equation}
where $ (s,t) $ denotes the ordered pair with $ s $ in the first position and $ t $ in the second; i.e. $ (s,t) = (t,s) $ if and only if $ s = t $.

A \df{function} $ f $ with \df{domain} $ D $ and \df{codomain} $ C $ (or a function \df{from} $ D $ \df{(in)to} $ C $) is a subset
of $ D \times T $ such that if $ (x,y_1) \in f $ and $ (x, y_2) \in f $ then $ y_1 = y_2 $. When defining such a function, we can
write that $ f : D \to C $. If $ (x,y) \in f $, we usually write $ f(x) = y $.

The set $ R = \{r \in C : f(x) = r \text{ for some } x \in D\} $ is called the \df{range} of $ f $. If $ R = C $ (that is, for every value $ y $
in the codomain there is some $ x $ in the domain so that $ f(x) = y $), then the function is called \df{surjective} or \df{onto}.

If $ f(x_1) = y $ and $ f(x_2) = y $ implies that $ x_1 = x_2 $ (that is, if every member $ y $ of the range has precisely one corresponding
element $ x $ of the domain so that $ f(x) = y $) then the function is called \df{injective} or \df{one-to-one}.

Functions which are both one-to-one and onto are called \df{bijections}; if there exists a bijection between two sets then those
sets are said to have the same \df{cardinality}. If a (finite) set $ S $ has a particular number of elements $ n $, then that number is also
called its cardinality (since there is a bijection between the set of natural numbers up to $ n $ and the elements of the set); we write $ \abs{S} = n $.

We say that $ f : A \to B $ and $ g : C \to D $ are \df{equal}, writing $ f = g $, if and only if both of the following conditions are met:
\begin{enumerate}
  \item $ A = C $ and $ B = D $; and
  \item For every $ x \in A $, $ f(x) = g(x) $.
\end{enumerate}

\begin{ex}\leavevmode
  \begin{enumerate}
    \item If $ f : \mathbb{R} \to \mathbb{R} $ is defined by $ f(x) = 3x + 1 $, then $ f $ is one-to-one and onto.
    \item If $ g : \mathbb{N} \to \mathbb{N} $ is defined by $ f(x) = x + 1 $, then $ f $ is one-to-one but not onto, since there is no $ x \in \mathbb{N} $
          such that $ 1 = x + 1 $.
  \end{enumerate}
\end{ex}

\begin{thm}\footnote{\df{Theorem}: A statement which has been proved to be true.}
  Let $ f : D \to C $ be a function, and let $ R \subseteq C $ be the range of $ f $. Then $ f $ is one-to-one and onto if and only if there exists a
  function $ g : C \to D $ so that for every $ x \in D $, $ x = g(f(x)) $ and for every $ y \in C $, $ y = f(g(y)) $. This function $ g $ is unique.
\end{thm}
\begin{proof}\footnote{\df{Proof}: A valid argument, from true premises, to obtain a conclusion. The symbol $ \qed $ denotes the end of a proof.}
  Suppose $ f $ is one-to-one and onto. Since $ f $ is one-to-one, we can define a new function $ g $ so that $ g(y) = x $ if and only if $ f(x) = y $:
  since if $ g(y) = x_1 $ and $ g(y) = x_2 $, we have $ f(x_1) = y = f(x_2) $ and $ x_1 = x_2 $ by the injectivity of $ f $. This new function $ g $
  has domain $ R = C $ (since $ f $ is onto, the two sets are equal) and range $ D $. In particular, if $ x \in D $, $ f(x) = y $ for some $ y \in R $
  and hence $ x = g(y) = g(f(x)) $. Similarly, if $ y \in R $, then there exists $ x \in D $ such that $ f(x) = y $ and so $ g(y) = x $; hence $ f(g(y)) = y $.

  Conversely, suppose such a $ g $ exists. Suppose $ f(x_1) = f(x_2) = y $ for some $ x_1, x_2 \in D $ and $ y \in R $. Then $ g(f(x_1)) = g(y) $, so $ g(y) = x_1 $.
  Similarly, $ g(f(x_2)) = g(y) $ so $ g(y) = x_2 $. Hence $ x_1 = x_2 $ and $ f $ is injective. Let $ y \in C $; then $ y = f(g(y)) $, and so there
  exists $ x = g(y) $ so that $ f(x) = y $; hence $ y \in R $, so $ C = R $ and $ f $ is onto.

  Suppose for uniqueness that both $ g_1 $ and $ g_2 $ satisfy the requirements of the theorem. Then for every $ y \in R $,  $ y = f(g_1(y)) = f(g_2(y)) $. But
  since $ f $ is injective, $ g_1(y) = g_2(y) $; and since $ g_1 $ and $ g_2 $ have the same domain and codomain, and agree at every point, they are equal.
\end{proof}

The unique function $ g $ defined as in the theorem for some function $ f $ is called the \df{inverse} of $ f $, and we usually write $ g = f^{-1} $. If
a function has an inverse then it is called \df{invertible}, and thus the theorem above shows that $ f $ is invertible if and only if it is a bijection.

\begin{exercise}
  Show that, if $ f^{-1} $ is the inverse of $ f $, then $ f^{-1} $ has an inverse and that inverse is $ f $.
\end{exercise}

\section{Combinatorics}
\subsection{Classes of Maps}
We will now look at combinatorics as an application of set theory.
\begin{warning}
  This section will not be nice and cuddly like the combinatorics we did last year; instead of counting players on a team or probabilities, we
  will be working directly with the guts of sets and functions. While this section is optional (like the rest of the notes), sometimes the gory
  detail can provide a bit of perspective and some clarity.
\end{warning}

The basic combinatorial rules, which are restatements of results from set theory, are:
\begin{equation}
  \begin{minipage}{0.8\textwidth}
    \begin{enumerate}
      \item \df{The rule of equality.} If $ N $ and $ R $ are finite sets, and if there exists a bijection between them, then $ \abs{N} = \abs{R} $.
      \item \df{The rule of sums.} If $ A_1, ..., A_n $ are finite sets with no elements in common, then $ \abs{\cup A_i} = \abs{A_1} + \cdots + \abs{A_n} $.
      \item \df{The rule of products.} If $ A_1, ..., A_n $ are finite sets, then $ \abs{A_1 \times \cdots \times A_n} = \abs{A_1} \cdots \abs{A_n} $.
    \end{enumerate}
  \end{minipage}
\end{equation}

For the remainder of this section, we will let $ N $ and $ R $ be finite sets. We will study the following sets of functions from $ N $ to $ R $, and
count their elements:
\begin{equation}
  \begin{aligned}
    \Map(N,R) &= \{f : N \to R : f \text{ arbitrary}\}\\
    \Sur(N,R) &= \{f : N \to R : f \text{ surjective}\}\\
    \Inj(N,R) &= \{f : N \to R : f \text{ injective}\}\\
    \Bij(N,R) &= \{f : N \to R : f \text{ bijective}\}
  \end{aligned}
\end{equation}

As a first step, we can say the following things about the sizes of $ N $ and $ R $ based on the existence (or lack thereof) of functions
matching these criteria.
\begin{lem}\leavevmode
  \begin{enumerate}
    \item If $ \Bij(N,R) $ is not empty, then $ \abs{N} = \abs{R} $.
    \item If $ \Inj(N,R) $ is not empty, then $ \abs{N} \leq \abs{R} $.
    \item If $ \Sur(N,R) $ is not empty, then $ \abs{N} \geq \abs{R} $.
  \end{enumerate}
\end{lem}
\begin{proof}
  If there is a bijection from $ N $ to $ R $, then we can pair up the elements of both sets with none left over, and hence the
  two sets are the same size (this is just the rule of equality). If there is an injective function from $ N $ to $ R $, then for
  every distinct element of $ N $ there is at least one distinct element of $ R $, so there are at least as many elements in $ R $
  as in $ N $. If there is a surjective function $ f $ from $ N $ to $ R $, then for every $ r \in R $ there is some element $ n \in N $
  so that $ f(n) = r $ and hence there are at least as many elements in $ N $ as there are in $ R $.
\end{proof}

\subsection{Orderings}
Suppose that we order $ N $ in some way, satisfying the following criteria:
\begin{enumerate}
  \item If $ a \leq b $ and $ b \leq a $, then $ a = b $.
  \item If $ a \leq b $ and $ b \leq c $, then $ a \leq c $.
  \item For all $ a $ and $ b $, either $ a \leq b $ or $ b \leq a $.
\end{enumerate}
(Such an ordering is called a \df{total order}.) Then we can look at any $ f : N \to R $ as giving a sequence of symbols in a
word: if $ n \in N $ then we label the place $ n $ in the word with the symbol $ s = f(n) $; another metaphor is that we label
a sequence of boxes with the ordered elements of $ N $ and then place the element $ f(n) $ of $ R $ in the corresponding box.
Thus any map $ f : N \to R $, if $ N $ is ordered, is a labelling, indexing, or addressing of elements of $ S $ with elements of $ N $.
If $ N = \{n_1, ..., n_m\} $ (with the natural total ordering), then we can unambigously represent $ f $ by writing the word $ f(n_1) f(n_2) \cdots f(n_m) $
(here we are not multiplying, simply writing each symbol next to its neighbours).

\begin{rem}
  If $ N $ is totally ordered but $ R $ is unordered, and we pick a function $ f : N \to R $, then the sets $ f(X) = \{f(x) : x \in X\} $ are
  all the subsets of $ R $ (with the usual conventions, i.e. $ \{a, a\} = \{a\} $ and $ \{a,b\} = \{b,a\} $). If we order $ R $, then we obtain
  the ordered multisubsets of $ R $ (i.e. ordered subsets of $ R $ that can contain multiple copies of each element).
\end{rem}

\begin{ex}
  Let $ N = \{1 < 2 < 3\} $ and $ R = \{a < b < c\} $. Then we can write out all of the elements of $ \Map(N,R) $:
  \begin{displaymath}
  \begin{tabular}{cccccc}
    abc & acb & bac & bca & cab & cba\\
    aab & aba & baa & aac & aca & caa\\
    bba & bab & abb & bbc & bcb & cbb\\
    cca & cac & acc & ccb & cbc & bcc\\
    aaa & bbb & ccc &     &     &
  \end{tabular}
  \end{displaymath}
  Then the first line consists of all the surjections; no other maps are onto or one-to-one.
\end{ex}

\begin{exercise}
  Show that, if $ N $ is finite, and there exists a bijection from $ N $ to $ R $, then all injective or surjective functions
  from $ N $ to $ R $ are in fact bijective.
\end{exercise}

\subsection{Permutations}
If $ N = R $, then any bijective function $ f : N \to R $ is called a \df{permutation} of $ N $ --- one can imagine that $ f $ acts
upon $ N $ by rearranging the order of its elements.

\section{Divisibility}
\subsection{The Definition}
Let $ a $ and $ b $ be integers. We say that $ a $ \df{divides} $ b $ (or $ a $ is a \df{factor} of $ b $) if there is some integer $ q $ such
that $ b = aq $; we write $ a \mid b $. If $ a $ does not divide $ b $, we write $ a \nmid b $.
\begin{lem}\footnote{\df{Lemma}: A statement which is proved on the way to a more important result. (Plural: lemmata.)}\leavevmode
  Let $ a $, $ b $, and $ c $ be integers. Then:
  \begin{enumerate}
    \item $ a \mid 0 $.
    \item If $ a \mid b $, and $ b \mid c $, then $ a \mid c $.
    \item If $ a \mid b $, and $ a \mid c $, then $ a \mid (b + c) $.
    \item If $ a \mid b $, then $ a \mid bc $.
%     \item If $ a \mid b $, and $ b \mid a $, then $ a = \pm b $.
  \end{enumerate}
\end{lem}
\begin{proof}\leavevmode
  \begin{enumerate}
    \item We must find an integer $ q $ so that $ 0 = qa $. Let us simply take $ q = 0 $; then $ 0 = 0a $ for all $ a $, so $ a \mid 0 $.
    \item Because $ a \mid b $, there exists some integer $ q $ so that $ b = qa $. Similarly, there exists some integer $ r $ so that $ c = rb $.
          Putting these together, $ c = rb = r(qa) = (rq)a $; so $ a \mid c $.
    \item Because $ a \mid b $, there exists some integer $ q $ so that $ b = qa $. Similarly, there exists some integer $ r $ so that $ c = ra $.
          Adding these, we have that $ b + c = qa + ra = (q + r)a $, and so $ a \mid (b+c) $.
    \item Because $ a \mid b $, there exists some integer $ q $ so that $ b = qa $. Hence $ bc = c(qa) = (cq)a $ and so $ a \mid bc $.
%     \item We have $ q $ and $ r $ so that $ b = qa $ and $ a = rb $. Hence $ b = qrb $, and so $ qr = 1 $. Then $ q = \pm 1 $, so $ b = \pm a $.
  \end{enumerate}
\end{proof}

From this point, lower case italic letters will always denote integers unless otherwise stated.

\begin{cor}\footnote{\df{Corollary}: A proof which is a direct result of an earlier proof.}
  If $ d \mid a_1 $, $ d \mid a_2 $, ..., $ d \mid a_n $, then
  \begin{displaymath}
    d \mid (c_1 a_1 + c_2 a_2 + \cdots c_n a_n)
  \end{displaymath}
  for any integers $ c_1 $, ..., $ c_n $.
\end{cor}
\begin{proof}
  By the definition of divisibility, there exist integers $ q_1 $, ..., $ q_n $ such that $ a_1 = q_1 d$, ..., $ a_n = q_n d $. Hence
  \begin{displaymath}
    c_1 a_1 + \cdots + c_n a_n = c_1 q_1 d + \cdots + c_n q_n d = d(c_1 q_1 + \cdots + c_n q_n)
  \end{displaymath}
  and so the result holds.
\end{proof}

\begin{exercise}
  Show that if $ a \mid b $, and $ a $ and $ b $ are positive, then $ a \leq b $. Show that this result
  does not always hold if $ a $ and $ b $ are non-negative.
\end{exercise}

\begin{exercise}
  If $ n $ is an integer, define the \df{absolute value} of $ n $ by
  \begin{displaymath}
    \abs{n} = \begin{cases}
                n & \text{if } n \geq 0\\
                -n & \text{if } n < 0.
              \end{cases}
  \end{displaymath}
  Show that $ \abs{n} \mid n $ and $ n \mid \abs{n} $.
\end{exercise}

\subsection{Greatest Common Divisors}
\begin{defn}
  If $ a $ and $ b $ are integers, then the integer $ d $ is called the \df{greatest common divisor} of $ a $ and $ b $ (written $ d = (a,b) $)
  if the following two conditions are met:
  \begin{enumerate}
    \item $ d \mid a $ and $ d \mid b $ (that is, $ d $ is a common divisor of both numbers).
    \item If $ d' \mid a $ and $ d' \mid b $, then $ d' \leq d $ (that is, every other common divisor is less than $ d $).
  \end{enumerate}
\end{defn}

\begin{ex}\leavevmode\label{ex:gcd}
  \begin{enumerate}
    \item Since for every $ a $ and $ b $ we have $ 1 \mid a $ and $ 1 \mid b $, it follows that $ (a,b) \geq 1 $.
    \item The greatest common divisor of $ -3 $ and $ 5 $ is $ (-3, 5) = 1 $, since $ 1 \mid -3 $ and $ 1 \mid 5 $ but no integer larger than $ 1 $ divides both.
    \item We also have $ (4, 14) = 2 $ and $ (24, 36) = 12 $.
    \item For all integers $ n $, $ (n, 1) = 1 $ (since the only integers dividing 1 are $ \pm 1 $, and $ 1 > -1 $).
    \item For all integers $ n $, $ (n, 0) = \abs{n} $ (since $ \abs{n} \mid 0 $ and $ \abs{n} \mid n $, so $ (n,0) \geq \abs{n} $; and no divisors
          of $ n $ are greater than $ \abs{n} $, so $ (n, 0) \leq \abs{n} $).
  \end{enumerate}
\end{ex}

\begin{exercise}
  Let $ d $ be an integer. Prove that if $ d $ is a positive integer, then $ (d, nd) = d $; and if $ d $ is negative, then $ (d, nd) = -d $.
\end{exercise}

If $ (a,b) = 1 $, then $ a $ and $ b $ are called \df{coprime} or \df{relatively prime}.

\begin{thm}
  If $ (a,b) = d $, then $ a/d $ and $ b/d $ are coprime.
\end{thm}
This theorem is intuitive: if we divide two numbers out by their greatest common divisor, they no longer have any common divisors greater than 1. This
observation essentially proves itself.
\begin{proof}
  We must show that $ (a/d, b/d) = 1 $. Let $ (a/d, b/d) = c $. Then $ c \mid a/d $ and $ c \mid b/d $; so there exist $ q_1 $ and $ q_2 $ such
  that $ a/d = cq_1 $ and $ b/d = cq_2 $; hence $ a = cq_1 d $ and $ b = c q_2 d $. This implies that $ cd $ is a common divisor of $ a $ and $ b $,
  and so $ cd \leq d $. Since $ d = (a,b) \geq 1 $ (by example \ref{ex:gcd}.1 above), we can divide through by $ d $ and obtain $ c \leq 1 $. Applying
  the same example again, since $ c $ is itself a greatest common divisor by definition, we have $ c \geq 1 $. Combining these two inequalities, $ c = 1 $.
\end{proof}

Our next goal is to find an easy (well, easier) way to compute the greatest common divisor of two numbers. To do this, we must develop the
idea of division with remainder like we learned in primary school. The main theorem which ensures that we can divide properly in the integers
is the following
\begin{thm}[The division algorithm]
  Let $ a $ and $ b \neq 0 $ be positive integers. Then there exist \emph{unique} integers $ q $ and $ r $, so that $ 0 \leq r < b $,
  and so that $ a = qb + r $. The integers $ q $ and $ r $ are respectively called the \df{quotient} and \df{remainder}.
\end{thm}

For the proof of this, we must use the \df{well-ordering principle}:
\begin{equation}
  \begin{minipage}{0.8\textwidth}
    If $ S $ is a non-empty set of integers which is bounded below (i.e. there is some $ b \in \mathbb{Z} $ so that
    if $ n \in S $ then $ n \geq b $), then $ S $ has a smallest element.
  \end{minipage}
\end{equation}


\begin{proof}[Proof of the division algorithm]
  Consider the set $ S $ defined by
  \begin{equation}
    S = \{a - nb : n \in \mathbb{Z} \} \cap \mathbb{N}_0.
  \end{equation}
  This is just the set of all non-negative numbers which can be written in the form $ a - nb $, for some integer $ n $. Since $ a \in S $, the
  set is non-empty; and since every $ x \in S $ satisfies $ x \geq 0 $, $ x $ is bounded below by 0. Hence $ S $ has a least element, $ a - qb $.
  This element is non-negative by definition, and must be less than $ b $: if it were larger than $ b $, we could consider $ a - (q + 1)b = (a - qb) - b $,
  which is in $ S $ and less than $ a - qb $, which is impossible since $ a - qb $ is the least element in $ S $. So if we let $ r = a - qb $,
  then $ 0 \leq r < q $, and $ a = ab + r $. So suitable $ q $ and $ r $ exist.

  For uniqueness, suppose that $ q_1 $, $ r_1 $ and $ q_2 $, $ r_2 $ both satisfy the conclusion of the theorem. Hence $ a = bq_1 + r_1 = b q_2 + r_2 $,
  and so
  \begin{equation}\label{eqn:divn}
    0 = b(q_1 - q_2) + (r_1 - r_2).
  \end{equation}
  Since $ b \mid 0 $ and $ b \mid b(q_1 - q_2) $, it follows that $ b $ divides $ r_1 - r_2 $. But we have $ 0 \leq r_1 < b $ and $ 0 \geq -r_2 > -b $,
  so $ -b < r_1 - r_2 < b $. The only multiple of $ b $ in between $ -b $ and $ b $ is zero, so $ r_1 = r_2 $ and hence (by \ref{eqn:divn}) $ q_1 = q_2 $.
\end{proof}

In order to compute the greatest common divisor easily, we will create a chain of integers so that the pairs of integers next to each other
share the same GCD. The division algorithm allows us to do this, because of the following lemma.

\begin{lem}\label{lem:euclidalg}
  If $ a = bq + r $, then $ (a,b) = (b,r) $.
\end{lem}
\begin{proof}
  By the hypothesis, $ (a,b) \mid r $, and so $ (a,b) $ is a common divisor of $ b $ and $ r $. Hence $ (a,b) \leq (b,r) $. On the other hand, $ (b,r) \mid a $
  (since it divides both numbers on the right-hand side of the equality in the hypothesis) and so $ (b,r) \leq (a,b) $ (since $ (b,r) $ is a common divisor
  of $ a $ and $ b $). Combining both inequalities, $ (b,r) = (a,b) $.
\end{proof}

Finally, we have an algorithm for computing the greatest common divisor.
\begin{thm}[The Euclidean algorithm]
  If $ a $ and $ b \neq 0 $ are positive integers, and
  \begin{align*}
    a &= bq_1 + r_1,\\
    b &= r_1q_2 + r_2,\\
    r_1 &= r_2 q_3 + r_3,\\\
    &\enspace \vdots\\
    r_k &= r_{k + 1} q_{k + 2} + r_{k + 2}
  \end{align*}
  where each $ r_{k + 2} $ is bounded like $ 0 \leq r_{k + 2} < r_{k + 1} $, then for $ k $ large enough
  we have $ r_{k - 1} = r_{k} q_{k + 1} $ (i.e. $ r_{k + 1} = 0 $) and $ (a,b) = r_k $.
\end{thm}
\begin{proof}
  Since each $ r_k $ is non-negative, and each $ r_k $ is less than $ r_{k - 1} $, the sequence $ b > r_1 > \cdots > r_k $
  has a least non-negative element, $ r_k $, which must be zero or we would be able to divide again by the division algorithm.
  Then $ r_{k - 1} = r_{k} q_{k + 1} $, and (by repeated application of lemma \ref{lem:euclidalg}) we have
  \begin{displaymath}
    (a,b) = (b,r_1) = (r_1, r_2) = \cdots = (r_{k - 1}, r_{k}) = (r_{k}, 0) = r_k.
  \end{displaymath}
\end{proof}

\begin{ex}
  Let us compute $ (240, 373) $. We have:
  \begin{align*}
    373 &= 1 \times 240 + 133\\
    240 &= 1 \times 133 + 107\\
    133 &= 1 \times 107 + 26\\
    107 &= 4 \times 26 + 3\\
    26 &= 8 \times 3 + 2\\
    3 &= 1 \times 2 + 1\\
    2 &= 1 \times 1 + 1\\
    1 &= {\color{red}1} \times 1.
  \end{align*}
\end{ex}

\begin{ex}
  Let us compute $ (120, 340) $. We have:
  \begin{align*}
    340 &= 2 \times 120 + 100\\
    120 &= 1 \times 100 + 20\\
    100 &= {\color{red}5} \times 20.
  \end{align*}
\end{ex}

\begin{cor}[B\'ezout's Lemma]
  If $ a $ and $ b $ are positive integers, then there exist integers $ x $ and $ y $ such that $ ax + by = (a,b) $.
\end{cor}
\begin{proof}
  By the proof of the Euclidean algorithm, for some $ k $ we have
  \begin{align*}
    a &= bq_1 + r_1,\\
    b &= r_1q_2 + r_2,\\
    r_1 &= r_2 q_3 + r_3,\\\
    &\enspace \vdots\\
    r_{k - 2} &= r_{k - 1} q_k + r_k,\\
    r_{k - 1} &= r_k q_{k + 1},
  \end{align*}
  and $ r_k = (a,b) $. Rearranging:
  \begin{align*}
    a - bq_1 &= r_1,\\
    b - r_1q_2 &= r_2,\\
    r_1 - r_2 q_3 &= r_3,\\\
    &\enspace \vdots\\
    r_{k - 2} - r_{k - 1} q_k &= r_k = (a,b),
  \end{align*}
  and it is clear that we can substitute each equation into the one below it, eliminating every $ r_i $ until
  we (in the final line) arrive at some equation of the form $ ax + by = (a,b) $.
\end{proof}

B\'ezout's Lemma is surprisingly useful; in fact, we will use it later this year to prove various properties of a particular system
in algebra.\footnote{For those following along from the future, it will be used when we study \df{primitive roots}.}

Since we know that $ (a,b) $ divides $ ax + by $ for all $ x $ and $ y $, the lemma tells us that for every $ n $, the equation
\begin{equation}
  ax + by = n(a,b)
\end{equation}
has an integer solution for $ x $ and $ y $. In fact, it has infinitely many:
\begin{thm}
  If $ (x,y) $ solves $ ax + by = z $, then all solutions are of the form $ (x - bn, an + y) $ for some $ n \in \mathbb{Z} $.
\end{thm}

\begin{proof}
  Clearly $ a(x - bn) + b(an + y) = z $, so all expressions of that form are indeed solutions. On the other hand, suppose $ (x', y') $
  is a solution. Then $ 0 = a(x - x') + b(y - y') $; it follows that $ a \mid (y - y') $. Rearranging, we obtain $ x' = x - b\frac{(y' - y)}{a} $,
  where $ (y' - y)/a = n $ is an integer. Hence $ z = a(x - bn) + by' = ax - abn + by' = ax + b(y' - an) $, and hence $ y' = y + an $.
\end{proof}

\begin{ex}
  The equation $ 18x + 24y = 23 $ has no solutions, since $ (18,24) = 6 $ does not divide $ 23 $.
\end{ex}
\begin{ex}
  We will solve $ 18x + 24y = 36 $ for all the integer solutions $ x, y $. Since $ (18,24) = 6 $, we will begin by solving $ 18x + 24y = 6 $:
  \begin{align*}
    24 = 1 \times 18 + 6\\
    18 = 6 \times 3
  \end{align*}
  Hence $ 6 = 1 \times 24 + (-1) \times 18 $, and $ 36 = 6 \times 24 + (-6) \times 18 $. By the theorem above, all the solutions are
  given by $ (6 - 24n, -6 + 18n) $ for $ n \in \mathbb{Z} $.
\end{ex}

\section{Induction}
Recall the well-ordering principle:
\begin{equation}
  \begin{minipage}{0.8\textwidth}
    If $ S $ is a non-empty set of integers which is bounded below (i.e. there is some $ b \in \mathbb{Z} $ so that
    if $ n \in S $ then $ n \geq b $), then $ S $ has a smallest element.
  \end{minipage}
\end{equation}

We will use this principle to prove the following \df{induction principle}.
\begin{thm}
  Let $ S \subseteq \mathbb{N} $ be a set satisfying the following criteria:
  \begin{enumerate}
    \item $ 1 \in S $
    \item If $ n \in S $, then $ n + 1 \in S $.
  \end{enumerate}
  Then $ S = \mathbb{N} $.
\end{thm}
\begin{proof}
  Let $ Q $ be the set of all natural numbers not in $ S $. Since every element of $ Q $ is greater than 1, $ Q $ is a bounded below subset of the
  integers and hence either is empty or has a least element, $ n $. Suppose the latter. Since $ 1 \notin Q $, $ n \neq 1 $. Hence $ n - 1 $ is a
  natural number, and must therefore be in $ S $ (or otherwise $ n $ would not be the least element of $ Q $). But, by the definition of $ S $,
  if $ n - 1 \in S $ then $ n = (n - 1) + 1 \in S $. Hence $ n \in S $, and $ Q $ does not have a least element; so it is empty, and $ S = \mathbb{N} $.
\end{proof}

This principle is frequently useful if we want to prove a statement about all natural numbers, but we don't have much information
to go on. We normally split proofs by induction into three steps:
\begin{enumerate}
  \item The base case. Prove that $ 1 \in S $, where $ S $ is the set of all natural numbers for which the property holds.
  \item The induction step. Prove that if the property holds for $ n $, then it holds for $ n + 1 $.
  \item The conclusion step. Apply the principle of induction to (1) and (2) above, and conclude that $ S = \mathbb{N} $: the property holds for all natural numbers.
\end{enumerate}

\begin{thm}
  The expression $ 6^n - 1 $ is divisible by 5 for all $ n \in \mathbb{N} $.
\end{thm}
\begin{proof}
  We use induction on $ n $.

  \paragraph{Base case: $ n = 1 $.}
  If $ n = 1 $, then $ 6^n - 1 = 6 - 1 = 5 $, which is obviously divisible by 5.

  \paragraph{Inductive step.}
  Suppose that the theorem holds for some $ n \in \mathbb{N} $. Consider $ 6^{n + 1} - 1 = 6 \cdot 6^n - 1 = 6 \cdot 6^n - 6 + 5 = 6(6^n - 1) + 5 $;
  since $ 6^n - 1 $ is divisible by 5, it follows that $ 6(6^n - 1) + 5 $ is divisible by 5 (why?) and so the theorem holds for $ n + 1 $.

  \paragraph{Conclusion.}
  Since the theorem holds for $ n = 1 $, and the result for $ n $ implies the result for $ n + 1 $, the theorem holds for all natural numbers.
\end{proof}

As a brief notational note, since we need to write sums and products of a series of numbers quite often, we introduce a condensed form:
\begin{gather}
  \sum_{k = 1}^n a_k = a_1 + a_2 + \cdots + a_n\\
  \prod_{k = 1}^n a_k = a_1 a_2 \cdots a_n.
\end{gather}

\begin{exercise}\label{ex:w1}
  Show that the sum of all the natural numbers, $ \sum_{k = 1}^n k $, is equal to $ \frac{n(n+1)}{2} $ using induction.
\end{exercise}

\begin{exercise}
  Find a closed-form formula for $ \sum_{k = 0}^n k^2 $.
\end{exercise}

Recall also that the \df{factorial} of $ n $ is defined by
\begin{displaymath}
  n! = 1 \cdot 2 \cdots (n - 1) \cdot n.
\end{displaymath}

\begin{exercise}
  Show that for all $ n \in \mathbb{N} $, $ 2^{n - 1} \leq n! $.
\end{exercise}

\begin{exercise}
  Use induction to show that, for all $ n $, $ x^n - 1 = (x - 1)\left(\sum_{k = 1}^{n - 1} x^k \right) $.
\end{exercise}

\begin{exercise}
  The set $ \mathcal{P}(X) $, called the \textbf{power set} of a set $ X $, is the set of all subsets of $ X $. Show
  that $ \abs{\mathcal{P}(X)} = 2^{\thinspace\abs{X}} $ if $ X $ is finite.
\end{exercise}

Induction is often useful when we are trying to prove statement about recursive definitions: definitions where
the $n$th case depends on the $ (n - 1)$th case.

\begin{exercise}
  Consider the sequence $ T $ defined recursively by
  \begin{align*}
    T_0 &= 0\\
    T_n &= 2T_{n - 1} + 1.
  \end{align*}
  For example, $ T_1 = 2 \times 0 + 1 = 1 $; $ T_2 = 2 \times 1 + 1 = 3 $; and $ T_3 = 2 \times 3 + 1 = 7 $. Can you find
  a closed-form formula for $ T_n $ (i.e. a formula depending only on $ n $)?
\end{exercise}

\begin{exercise}
  Recall that the \df{Fibonnacci series} is defined by
  \begin{align*}
    F_1 &= F_2 = 1\\
    F_n &= F_{n - 1} + F_{n - 2}.
  \end{align*}
  The first few values of this sequence are $ 1, 1, 2, 3, 5, 8, 13, \dots $. At first glance, there does not seem to be
  any simple closed-form equation for $ F_n $; it is surprising, then, that
  \begin{displaymath}
    F_n = \frac{\left(1 + \sqrt{5}\right)^n - \left(1 - \sqrt{5}\right)^n}{2^n \sqrt{5}}.
  \end{displaymath}
  Prove it.
\end{exercise}


\section{Prime Factorisation}
\begin{defn}
  A \df{prime} is a number $ p \in \mathbb{N} $ such that if there exists $ n \in \mathbb{N} $ where $ n \mid p $, then $ n = 1 $ or $ n = p $.
\end{defn}

The primes are, in some sense, the `building blocks' of the natural numbers. This is because every natural number can
be written as a product of prime numbers, and be written like that in exactly one way. Our goal is, eventually, the
following theorem which guarantees this property; we will give the proof at the end of the section.
\begin{thm}[Fundamental Theorem of Arithmetic]\label{thm:funarithmetic}
  If $ n $ is a natural number, then there exist distinct primes $ p_1 < p_2 < \cdots < p_n $ and natural
  numbers $ e_1 $, $ e_2 $, ..., $ e_n $, such that
  \begin{displaymath}
    n = p_1^{e_1} p_2^{e_2} \cdots p_n^{e_n}.
  \end{displaymath}
  This factorisation is unique.
\end{thm}

\begin{ex}
  We can write $ 1234 = 2 \times 617 $, and $ 540 = 2^2 \times 3^3 \times 5 $.
\end{ex}

We start out with some properties of the primes which we take for granted.
\begin{lem}[Euclid's lemma]
  If $ p $ is prime, and $ p \mid ab $, then either $ p \mid a $ or $ p \mid b $.
\end{lem}
Clearly this is not necessarily true if $ p $ is not prime:- $ 6 \mid (3 \times 4) $, but $ 6 \nmid 3 $ and $ 6 \nmid 4 $!
\begin{proof}
  Suppose $ p \mid ab $. Consider $ (p,b) $; since it is positive and divides $ p $, it is equal to either 1
  or $ p $. If $ (p,b) = p $, then $ p \mid b $ and we are done. So suppose that $ (p,b) = 1 $. Then there exist (by B\'ezout's
  lemma) $ x $ and $ y $ such that $ xp + yb = 1 $; multiplying both sides by $ a $, $ apx + aby = a $. Since $ p $ divides both
  terms on the left, it must divide the right hand side; so $ p \mid a $, and we are done.
\end{proof}

\begin{lem}
  If $ n \neq 1 $ is a natural number, then there exists some prime $ p $ such that $ p \mid n $.
\end{lem}
\begin{proof}
  Let $ n $ be the smallest natural number not divisible by any prime. Then $ n $ is not itself a prime,
  and so there is some natural number $ m $, $ 1 < m < n $, that divides $ n $. But $ m $ is divisible by
  a prime $ p $ (since it is less than $ n $), and hence $ p \mid n $. So there cannot be a smallest natural
  number not divisible by any prime, and hence all natural numbers are divisible by primes (except 1).
\end{proof}

These two lemmas are all we need for the
\begin{proof}[Proof of the fundamental theorem of arithmetic]
  Since the theorem is obvious for $ n $ prime, let $ n > 1 $ be a non-prime natural number. We use induction; assume that all the $ m < n $ are uniquely
  factorisable into primes.
  \paragraph{Existence of factorisation.}
  Since $ n > 1 $, there is some prime $ p $ and integer $ m \leq n $ so that $ pm = n $. By the inductive hypothesis, $ m $ is factorisable into
  primes; so $ n $ is factorisable into primes.
  \paragraph{Uniqueness of factorisation.}
  Suppose that $ n $ can be factorised in two ways, as $ n = p_1^{e_1} \cdots p_n^{e_n} = q_1^{f_1} \cdots q_m^{f_m} $. By Euclid's lemma,
  we have $ p_1 = q_i $ for some $ i $; hence $ p_1^{e_1 - 1} \cdots p_n^{e_n} = q_1^{f_1} \cdots q_i^{f_i - 1} \cdots q_m^{f_m} $. But this
  new number is less than $ n $, and so the factorisation is unique; applying the inductive hypothesis, the factorisation of $ n $ is unique.
\end{proof}

We now look at a couple of applications of the fundamental theorem. First of all,
\begin{cor}
  If $ (a,b) $ and $ [a,b] $ are, respectively, the greatest common divisor and least common multiple of $ a $ and $ b $,
  then $ (a,b)[a,b] = ab $.
\end{cor}
\begin{proof}
  Suppose $ a $ and $ b $ can be factorised as
  \begin{gather*}
    a = p_1^{e_1} \cdots p_n^{e_n}\\
    b = p_1^{f_1} \cdots p_n^{f_n}
  \end{gather*}
  where we write both in terms of the same primes, but exponents may be zero. I claim that
  \begin{displaymath}
    (a,b) = p_1^{\min(e_1, f_1)} \cdots p_n^{\min(e_n, f_n)} \text{ and } [a,b] = p_1^{\max(e_1, f_1)} \cdots p_n^{\max(e_n, f_n)}
  \end{displaymath}
  (check this yourself), and by direct multiplication the result holds.
\end{proof}

\begin{cor}
  If $ n $ has prime factorisation $ n = p_1^{e_1} \cdots p_n^{e_n} $, then the number of possible factors of $ n $ is precisely $ d(n) = \prod_{k = 1}^n (e_k + 1) $.
\end{cor}
\begin{proof}
  By Euclid's lemma, every divisor of $ n $ is of the form $ p_1^{f_1} \cdots p_n^{f_n} $ where each $ f_k $ satisfies $ 0 \leq f_k \leq e_k $. There
  are $ e_k + 1 $ choices for each exponent, and each exponent is independent; the result follows by counting.
\end{proof}

Finally, we will look at one final classical application of Euclid's lemma:
\begin{cor}
  $ \sqrt{p} $ is irrational for all primes $ p $.
\end{cor}
\begin{proof}
  Suppose $ \sqrt{p} $ is rational; so there exist coprime integers $ a $ and $ b $ such that $ \sqrt{p} = a/b $. This implies
  that $ pb^2 = a^2 $. Since $ p \mid a^2 $, by Euclid's lemma $ p \mid a $; hence $ a = np $ for some $ n $, and $ pb^2 = n^2 p^2 $.
  Cancelling $ p $, $ b^2 = n^2 p $, and (again by Euclid's lemma) $ p \mid b $. So $ p \leq (a,b) $, and (in particular) $ (a,b) > 1 $ ---
  so the existence of fractional representation in lowest form of $ \sqrt{p} $ is contradictory.
\end{proof}

\section{Fields}

\section{Next Steps}
As well as the material in the level three standards (differentiation, integration, algebra, linear systems, linear programming, path analysis,
trigonometry, and conic sections) it is advantageous for Scholarship students to be aware of mathematics as a broader discipline. In the 2017
paper, for example, the opening question part was on number theory.

As such, I have here a list of books and topics which Scholarship students may wish to look over. I have included the University of Auckland
library call number in brackets.
\begin{itemize}
  \item \textbf{Culture:} \emph{A Mathematician's Lament} by Paul Lockhart [510.71 L81].
  \item \textbf{Proofs:} \emph{How to Think Like a Mathematician} by Kevin Houston [510 H84].
  \item \textbf{Number theory:} \emph{Elementary Number Theory} by Underwood Dudley [512.72 D84].
  \item \textbf{Linear algebra:} \emph{Linear Algebra: A Modern Introduction} by David Poole [512.5 P82].
  \item \textbf{Geometry:} \emph{Geometry: A High School Course} by Serge Lang and Gene Murrow [516 L27].
  \item \textbf{Set theory:} \emph{Naive Set Theory} by Paul Halmos [511.322 H19].
  \item \textbf{Calculus:} \emph{Calculus} by Paul Spivak [515 S76].
\end{itemize}

I must stress that a comprehensive understanding of the content in the above books is \emph{not required}; in most cases, a read of the first three
sections will suffice to obtain an overview of the relevant parts of the subject (not that that precludes interested students from reading further).

\end{document}
