\input{stdinclude.tex}
\begin{document}

\mainHeaderDiffHw{5}{The Product and Quotient Rules}
\subsection*{Reading}
Even next year, when you will be expected to know the derivative rules, it is not necessary to remember the quotient rule because
it is just a special case of the product rule. However, here is a little rhyme:

\begin{center}\itshape
  If it's the quotient rule you wish to know,\\
  It's low-de-high less high-de-low.\\
  Then draw the line and down below,\\
  Denominator squared will go.\footnote{Quoted in \textit{Mathematical Apocrypha} by Steven G. Krantz (p.36).}
\end{center}

\subsection*{Questions}
\begin{questions}
  \question Find the derivatives:
    \begin{parts}
      \part $ \od{y}{x} $ if $ y = \sin x \ln x $.
      \part $ \od{y}{x} $ if $ y = x \sec kx $ ($ k $ constant).
      \part $ \od{f}{\theta} $ if $ f(\theta) = \frac{\cos \pi \theta}{\sin \pi \theta + \cos \pi \theta} $.
      \part $ \od{y}{t} $ if $ y = \cos^4 (\sin^3 t) $.
    \end{parts}
  \question The force $ F $ acting on a body with mass $ m $ and velocity $ v $ is the rate
            of change of momentum, $ F = \od{}{t} [mv] $. If $ m $ is constant, this becomes $ F = ma $,
            where $ a = \od{v}{t} $ is the acceleration of the body. However, due to relativistic
            effects, the mass of a particle varies with $ v $ as
            \begin{displaymath}
              m = \frac{m_0}{\sqrt{1 - \dfrac{v^2}{c^2}}},
            \end{displaymath}
            where $ m_0 $ is the rest mass of the body and $ c $ is the speed of light. Show that
            \begin{displaymath}
              F = \frac{m_0 a}{\left(1 - \dfrac{v^2}{c^2}\right)^{\frac{3}{2}}}.
            \end{displaymath}
  \question Recall that if $ \theta $ is given in degrees, then $ \frac{\pi \theta}{180} $ is the equivalent angle in radians.
            Find the derivative of $ \sin \theta $ if $ \theta $ is given in degrees.
\end{questions}
\end{document}
