\input{stdinclude.tex}
\begin{document}

\mainHeaderDiffHw{11}{Related Rates of Change}
\subsection*{Reading}
\subsubsection*{Problem-Solving Strategy}
\begin{enumerate}
  \item Read the problem carefully.
  \item Identify the quantities given in the problem.
  \item Identify the unknown quantity.
  \item Draw a diagram and assign symbols to all quantities.
  \item Write an equation relative the various quantities together (if necessary, using
        substitution to eliminate unwanted variables).
  \item Use the chain rule to differentiate and solve for the unknown.
\end{enumerate}

\subsection*{Questions}
\begin{questions}
  \question If $ V $ is the volume of a cube with edge length $ x $ and the cube expands as time passes,
            find $ \od{V}{t} $ in terms of $ \od{x}{t} $.
  \question A water tank has the shape of an inverted circular cone with base radius \SI{2}{\metre}
            and height \SI{4}{\metre}. If water is being pumped into the tank at a rate of \SI{2}{\metre\cubed\per\minute},
            find the rate at which the water level is rising when the water is \SI{3}{\metre} deep.
  \question A boat is pulled into a dock by a rope attached to the bow of the boad and passing through a pulley on
            the dock that is \SI{1}{\metre} higher than the bow of the boat. If the rope is pulled in at a rate
            of \SI{1}{\metre\per\second}, how fast is the boat approaching the dock when it is \SI{8}{\metre} from the dock?
\end{questions}
\end{document}
